---
title: "Associations between Socioeconomic Factors and Breast Cancer Incidence in Ohio Counties"
author: "Olivia Ling and Jojo Yu"
date: "November 28, 2025"
output:
  html_document:
    toc: true
    toc_float: true
  pdf_document: default
---

## 0. Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

# IMPORTANT:
# Do NOT run install.packages() inside the Rmd.
# Instead, run this ONCE in the Console manually:
# install.packages(c("tidyverse","readxl","ggthemes","maps","mapdata","stringr"))

library(tidyverse)
library(readxl)
library(ggthemes)
library(maps)
library(mapdata)
library(stringr)
library(ggmap)
```
## 1. Read & clean data

```{r read_clean}
county_df <- read_excel("County data.xlsx", skip = 4) |>
  dplyr::rename(
    county      = `County`,
    rural_urban = `Rural/Urban`,
    app_flag    = `...3`          # third column (Appalachian info)
  ) |>
  dplyr::filter(!is.na(county)) |>
  dplyr::mutate(
    county = stringr::str_trim(county),
    rural_urban = factor(
      rural_urban,
      levels = c("Urban", "Partially Rural County", "Rural")
    ),
    # app_flag column is "Appalachian County" or NA
    app_flag = if_else(
      is.na(app_flag),
      "Non-Appalachian",
      "Appalachian"
    ),
    app_flag = factor(app_flag, levels = c("Non-Appalachian","Appalachian"))
  )

glimpse(county_df)
```


## 2. 
```{r}
ru_summary <- county_df |>
dplyr::count(rural_urban) |>
dplyr::mutate(pct = round(100 * n / sum(n), 1))
ru_summary

app_summary <- county_df |>
dplyr::count(app_flag) |>
dplyr::mutate(pct = round(100 * n / sum(n), 1))
app_summary

cross_tab <- county_df |>
dplyr::count(app_flag, rural_urban) |>
dplyr::group_by(app_flag) |>
dplyr::mutate(pct_within_app = round(100 * n / sum(n), 1))
cross_tab
```
## 3. Read other sheets & merge

```{r read_other_sheets}
# Each sheet: first column = County, second column = the variable we want.
# We use select(1, 2) so it doesn't matter what the original header text is.

poverty_df <- read_excel("County data.xlsx", sheet = "Poverty",   skip = 4) |>
  dplyr::select(1, 2) |>
  stats::setNames(c("county", "poverty_rate")) |>
  dplyr::filter(!is.na(county)) |>
  dplyr::mutate(county = stringr::str_trim(county))

income_df <- read_excel("County data.xlsx", sheet = "Income",    skip = 4) |>
  dplyr::select(1, 2) |>
  stats::setNames(c("county", "median_income")) |>
  dplyr::filter(!is.na(county)) |>
  dplyr::mutate(county = stringr::str_trim(county))

education_df <- read_excel("County data.xlsx", sheet = "Education", skip = 4) |>
  dplyr::select(1, 2) |>
  stats::setNames(c("county", "college_plus_pct")) |>
  dplyr::filter(!is.na(county)) |>
  dplyr::mutate(county = stringr::str_trim(county))

population_df <- read_excel("County data.xlsx", sheet = "Population", skip = 4) |>
  dplyr::select(1, 2) |>
  stats::setNames(c("county", "female_pop")) |>
  dplyr::filter(!is.na(county)) |>
  dplyr::mutate(county = stringr::str_trim(county))

cancer_df <- read_excel("County data.xlsx", sheet = "Cancer",   skip = 4) |>
  dplyr::select(1, 2) |>
  stats::setNames(c("county", "cancer_incidence")) |>
  dplyr::filter(!is.na(county)) |>
  dplyr::mutate(county = stringr::str_trim(county))

# Merge everything into one master dataset
full_df <- county_df |>
  dplyr::left_join(poverty_df,    by = "county") |>
  dplyr::left_join(income_df,     by = "county") |>
  dplyr::left_join(education_df,  by = "county") |>
  dplyr::left_join(population_df, by = "county") |>
  dplyr::left_join(cancer_df,     by = "county")

glimpse(full_df)
```
## 4. Quick summaries

```{r summaries_soc}
full_df |>
  dplyr::select(poverty_rate, median_income, college_plus_pct, female_pop, cancer_incidence) |>
  summary()
full_df$poverty_rate<-as.numeric(full_df$poverty_rate)
full_df$median_income<-as.numeric(full_df$median_income)
```


## 5. Graphs: Cancer incidence vs predictors

```{r plots_predictors, fig.width=6, fig.height=4}
# Poverty vs cancer
ggplot(full_df, aes(x = poverty_rate, y = cancer_incidence)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "Breast Cancer Incidence vs Poverty Rate",
    x = "Poverty rate (%)",
    y = "Breast cancer incidence (per 100,000)"
  ) +
  theme_minimal()

# Income vs cancer
ggplot(full_df, aes(x = median_income, y = cancer_incidence)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "Breast Cancer Incidence vs Median Income",
    x = "Median household income ($)",
    y = "Breast cancer incidence (per 100,000)"
  ) +
  theme_minimal()

# Education vs cancer
ggplot(full_df, aes(x = college_plus_pct, y = cancer_incidence)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "Breast Cancer Incidence vs % with College Degree+",
    x = "% adults with college degree or higher",
    y = "Breast cancer incidence (per 100,000)"
  ) +
  theme_minimal()
```

## 6. Use the median incidence as the cutoff for "high" vs "low"

```{r}
median_inc <- median(full_df$cancer_incidence, na.rm = TRUE)

full_df <- full_df |>
mutate(
high_incidence = if_else(
cancer_incidence > median_inc,
1L, 0L
),
high_incidence = factor(high_incidence, levels = c(0, 1))
)

# Check balance of the two groups

table(full_df$high_incidence)
```

## 11. Load Ohio county polygons

```{r}
ohio_map <- map_data("county") %>%
dplyr::filter(region == "ohio") %>%
dplyr::mutate(county = tolower(subregion))
```

## 12. Load Poverty data from County data.xlsx
# (sheet named exactly "Poverty"; header row is 4, so skip = 4)

```{r}
poverty <- read_excel(
  "County data.xlsx",
  sheet = "Poverty",
  skip = 4   # THIS is the correct skip
) %>%
  transmute(
    county = stringr::str_to_lower(stringr::str_remove(County, " County$")),
    poverty_pct = as.numeric(`Value (Percent)`)
  )

# quick checks (printed in knitted doc)
head(poverty)
head(ohio_map)
```


## 13. Merge map polygons with poverty data

```{r}
map_df <- ohio_map %>%
dplyr::left_join(poverty, by = "county") %>%
dplyr::arrange(order)
```

## 14. Compute centroids for county labels

```{r}
centroids <- map_df %>%
dplyr::group_by(county) %>%
dplyr::summarise(
clong = mean(long, na.rm = TRUE),
clat = mean(lat, na.rm = TRUE),
.groups = "drop"
)
```

## 15. Draw the heat map

```{r}
ggplot() +
geom_polygon(
data = map_df,
aes(x = long, y = lat, group = group, fill = poverty_pct),
color = "black",
linewidth = 0.2
) +
coord_fixed(1.3) +
geom_text(
data = centroids,
aes(x = clong, y = clat, label = county),
size = 2.0,
color = "black"
) +
scale_fill_distiller(palette = "Spectral", direction = 1) +
labs(
title = "Family Poverty (%) by County in Ohio",
fill = "Poverty %"
) +
theme_void() +
theme(
legend.position = "top",
plot.title = element_text(hjust = 0.5)
)
```
